---
title: "Coastal Ecosystem Accounting in the Main Hawaiian Islands"
subtitle: "Lorem ipsum odor amet, consectetuer adipiscing elit. Erat suspendisse proin massa eu suspendisse purus velit. Interdum finibus porttitor litora et facilisi cras. Pretium vel faucibus maecenas, pellentesque nec himenaeos nulla. Egestas sed pellentesque semper adipiscing netus arcu ut euismod. Aenean consequat quam posuere magna erat tortor mi phasellus? Finibus augue pretium tempus tempus leo mi; gravida congue lobortis."
format:
  lumo-html: 
    logo: "/assets/logos/uhm_logo.png"       
    # github-repo: "https://github.com/olesonlab/cea"        
    self-contained: true
    is-particlejs-enabled: false
    toc-title: "Contents"
author: 
  - name: Louis Chua
  - name: Ela Ural
  - name: Alemarie Ceria
  - name: Kirsten Oleson
date: last-modified
---

## What is Natural Capital Accounting (NCA)?

- A useful tool in measuring changes in stock and condition of natural capital
- ‘Accounts for’ the value of ecosystem services

## What's the Value in Valuing Nature?

Having baseline measures for the value of our ecosystems allows us to: 

- Form a more comprehensive calculation of GDP
- Have better information for use in cost-benefit analyses 
    - For example: What is the value of this fishery conservation policy? Will 
    the gains from fees on fishing justify the cost of the policy? 
- Creates data on natural resource flow for other sorts of analyses 

## Compiling Coastal Ecosystem Accounts (CEA) for Hawaii

Globally, despite being critical for food security, the economy, and protection 
from disasters, coasts face multiple threats, such as development, pollution, 
and over harvest. Hawaii is no exception. Coasts are threatened by sediment from 
former plantations, groundwater discharge laden with nutrients and pathogens, 
coastal development and hardening, and over use, to name a few. The resulting 
loss of valuable ecosystems and their services undermines human well-being and 
sustainability. We need a way to represent the importance of nature for the 
economy and human activity, so we make informed decisions and avoid unintended 
trade-offs. That is why we are compiling coastal ecosystem accounts for the state.

## Working Under the United Nation's (UN) System of Environmental Economic Accounting (SEEA-EA) Framework

- The UN has an existing platform
- Their statistical framework is standardized and familiar to policy-makers
- There are existing guidance and recommendations
- Tracks at relevant ecological spatial scales, changes in ecosystem extent, 
conditions, services, and asset value over time

## Accounting Spatial Unit: Main Hawaiian Islands (MHI)

The MHI is some of the most isolated islands on Earth. It has a resident 
population of approximately 1.4 million, 1 million of whom live on the island of 
Oahu (in Honolulu city and county). It also hosts over 10 million tourists a year 
(pre-COVID). Coastal ecosystems underpin tourism, provide food, and are sites for 
reflection and recreation. These ecosystems are critical to the economy, culture, 
and human well-being.

:::{.callout}
Our goal is to build accounts that span two periods, prior to and after the 
catastrophic coral reef bleaching event of 2014-2015
:::

## Accounting Units: Moku and Ecosystem Types (ET)

Ecosystem accounts are reported in a consistent spatial unit, which we define as 
the moku, a traditional Hawaiian resource management unit that roughly follows 
watershed boundaries. By extending from mountain out to the ocean, moku have the 
advantages of tying the land to the sea, and they resonate with managers and the 
public. 

Marine ET's are defined by the benthic zone. Our ET's include soft bottom, other 
hard bottom, pavement, and coral dominated soft bottom. Anything outside of these 
are defined as open ocean. Terrestrial ET's include developed, grass/shrub, tree 
cover, estuarine wetlands, barren, cropland, pasture, beaches/dunes, and 
freshwater wetlands.

## Extent Accounts

### Marine

Marine extent comes from HIMARC, a data collaborative led by Mary Donovan. Marine 
data layers represent the output of a spatially predictive model, and are based 
on data from the period 2004-2014. Ecologists are using these maps as the 
“pre-bleaching” extent.

### Terrestrial

## Conditions Accounts

### Marine 

### Terrestrial

## Services Accounts

### Fisheries

#### Physical Flow

*Marine Landings*

```{r, eval=FALSE, include=FALSE}
# Quantified in commercial and non-commercial fish landings
# 
# - Commercial data comes from the Hawai`i Department of Land and Resources (DLNR)'s 
# Division of Aquatic Resources (DAR) Commercial Marine Landings (CML) reportings 
#   - Fish Species Over Ecosystem Type (FSOET) Dataset
#     - Allocate physical flow in spatial terms over our ET's by assigning the 
#     probability of each species group to have been extracted from a particular 
#     ET
#     - Probabilities are based off of:
#       - Understanding of habitat: How long does a species spend in a particular 
#       ET?
#       - Migration knowledge: Where is a species known to travel to?
#       - Local fishing knowledge: In which habitat is the species best known to be 
#       caught? 
#     - We use FSOET’s ET probabilities to derive a distribution of catch for each 
#     species group over each ET
#   - Species groups include Deep 7 bottomfish, shallow bottomfish, pelagics, and
#   reef-associated
#   - Landings by DAR catch area
#   
# - Non-commercial data comes from the Marine Recreational Information Program 
# (MRIP)'s Hawai`i Marine Recreational Fishing Survey (HMRFS) 
#   - Focusing on reef herbivores, including kala (bluespine unicornfish), kole 
#   (goldring surgeonfish), manini (convict tang), spectacled parrotfish, redlip 
#   parrotfish, and uhu (all other parrotfish species)
#   - Landings by island
```

#### Monetary Flow

*Exchange Value (EV)*

```{r, eval=FALSE, include=FALSE}
# :::{.callout}
# SEEA-EA framework's definition of Exchange Value (EV): 
# 
# "[T]hose values that reflect the price at which ecosystem services and ecosystem 
# assets would be exchanged between buyer and seller if a market existed."
# 
# - United Nations (2017) 
# :::
# 
# ::: {.cell .page-columns .page-full .column-screen .grey-section}
# For commercial fish, we calculate net revenue. Making use of market prices is 
# "most preferred" method to estimate EV under SEEA-EA. Our valuation is suitable
# for a Net Present Value (NPV) approach (United Nations et al. 2014)
# 
# $$
# EV_{t} = \textit{landings} \times \textit{price} - \textit{Marginal Costs}_{t}
# $$
# :::
# 
# - For non-commercial fish, we assume that the value of fish caught to consume
# non-commercially is equivalent, by this "if a market existed" language, to that 
# commercial value
```

*Revenue*

## Benefit